---
title: "Class and Death: Workers Vs. Managers in Agriculture"
subtitle: "Exploration of Differences by Class in Agricultural Work"
author: "Luke Slipski"
date: "Published on `r format(Sys.time(), '%m-%d-%y')`"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 5
    page-layout: full
fig-cap-location: top
editor:
  markdown:
    wrap: 72
execute: 
  echo: false
  output: false
---

```{r}
library("tidyverse")
library("lubridate")
# download CAdeathserver with remotes::install_git("https://github.com/Lslipski/CAdeathserver")
library("CAdeathserver")
# download merced.clc package with remotes::install_git("https://github.com/Lslipski/merced.clc")
library("merced.clc")
# for working with ICD codes
# install with 
# remotes::install_local("C:\\Users\\luke_slipski\\Downloads\\icd-main.zip")
# after downloading to local from https://github.com/cran/icd
library("icd") 
```

This section will limit the cohort to only 2 occupation codes: 
```{r}
#| output: true
occ_codes_of_interest = tribble(~census_occ_code_2018, ~description,
                                "6050", "Miscellaneous Agricultural Workers",
                                "6040", "Graders and sorters, agricultural products",
                                "6005", "First-line supervisors of farming, fishing, and forestry workers",
                                "6010", "Agricultural inspectors",
                                "6020", "Animal breeders",
                                "6115", "Fishing and hunting workers",
                                "6120", "Forest and conservation workers",
                                "6130", "Logging workers",
                                "0205", "Farmers, ranchers, and other agricultural managers",
                                "7810", "Butchers and other meat, poultry, and fish processing workers")
cohort1_occ_codes = c("6050", "0205")
occ_codes_of_interest %>% 
  filter(census_occ_code_2018 %in% cohort1_occ_codes) %>% 
  dt_table(search = F,
           page = F)
```

```{r}
df_cohort1 = read_rds(here::here("data/df_cohort1.rds"))
```

```{r}
# original creation of cohort 1
# df_cohort1 = df_mortality_clean %>% 
#   filter(census_occ_code %in% cohort1_occ_codes)
```

```{r}
#|eval: FALSE
# save cohort 1 as rds for faster loading
# write_rds(df_cohort1,
#           file = here::here("data/df_cohort1.rds"))
```


## Number of Deaths

### Overall
```{r}
#| output: true
c1_num_deaths = df_cohort1 %>% 
  group_by(census_occ_title) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop") %>% 
  mutate(measure = "Number of Deaths",
         level = "Overall",
         percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                         " (",
                                                         percent_within_occupation,
                                                         "%)")) %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`)

c1_num_deaths %>% 
  dt_table() 
```

## Education

### Overall
```{r}
#| output: true
df_cohort1 %>% 
  count(recode_educ_farmwork) %>% 
  dt_table() %>% 
  add_commas("n")
```
### By Occupation
```{r}
#| output: true
c1_education = df_cohort1 %>% 
  group_by(census_occ_title, recode_educ_farmwork) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop_last") %>%
  mutate(percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                          " (",
                                                          percent_within_occupation,
                                                          "%)")) %>%
  ungroup() %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  rename(level = recode_educ_farmwork) %>% 
  mutate(measure = "Highest Level of Education") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_education %>% 
  dt_table(page = F) 
```












### By Occupation (Among those with Age >= 25)
```{r}
#| output: true
c1_education_age = df_cohort1 %>% 
  filter(recode_age_int >= 25) %>% 
  group_by(census_occ_title, recode_educ_farmwork) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop_last") %>%
  mutate(percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                          " (",
                                                          percent_within_occupation,
                                                          "%)")) %>%
  ungroup() %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  rename(level = recode_educ_farmwork) %>% 
  mutate(measure = "Highest Level of Education") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_education_age %>% 
  dt_table(page = F) 
```














## Marital Status

:::{.callout-note}
We may want to combine married/widowed/divorced with the domestic partnership versions of these.
:::

### Overall
```{r}
#| output: true
df_cohort1 %>% 
  count(recode_marital_status) %>% 
  dt_table() %>% 
  add_commas("n")
```


### By Occupation
```{r}
#| output: true
c1_marital_status = df_cohort1 %>% 
  group_by(census_occ_title, recode_marital_status) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop_last") %>%
  mutate(percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                          " (",
                                                          percent_within_occupation,
                                                          "%)")) %>%
  ungroup() %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  rename(level = recode_marital_status) %>% 
  mutate(measure = "Marital Status") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_marital_status %>% 
  dt_table(page = F) 
```

## Sex

### Overall
```{r}
#| output: true
df_cohort1 %>% 
  count(recode_sex) %>% 
  dt_table() %>% 
  add_commas("n")
```

### By Occupation
```{r}
#| output: true
c1_sex = df_cohort1 %>% 
  group_by(census_occ_title, recode_sex) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop_last") %>%
  mutate(percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                          " (",
                                                          percent_within_occupation,
                                                          "%)")) %>%
  ungroup() %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  rename(level = recode_sex) %>% 
  mutate(measure = "Sex") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_sex %>% 
  dt_table(page = F) 
```

## Race

:::{.callout-note}
We have a number of race codes that we still need to categorize. They're spelled out in the Class and Death Shared Notebook google doc
:::

### Overall
```{r}
#| output: true
df_cohort1 %>% 
  count(recode_race) %>% 
  dt_table() %>% 
  add_commas("n")
```


### By Occupation
```{r}
#| output: true
c1_race = df_cohort1 %>% 
  group_by(census_occ_title, recode_race) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop_last") %>%
  mutate(percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                          " (",
                                                          percent_within_occupation,
                                                          "%)")) %>%
  ungroup() %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  rename(level = recode_race) %>% 
  mutate(measure = "Race") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_race %>% 
  dt_table(page = F) 
```





## Age at Death


### Overall
```{r}
#| output: true
df_cohort1 %>% 
  summarize(mean_age_at_death = round(mean(recode_age_int, 
                                      na.rm = T), 1),
            std_age_at_death = round(sd(recode_age_int,
                                  na.rm = T), 1),
            .groups = "drop") %>% 
  mutate(`mean (sd)` = paste0(mean_age_at_death, " (", std_age_at_death, ")")) %>% 
  select(`mean (sd)`)
```


### By Occupation

```{r}
#| output: true
c1_age_at_death = df_cohort1 %>% 
  group_by(census_occ_title) %>% 
  summarize(mean_age_at_death = round(mean(recode_age_int, 
                                      na.rm = T), 1),
            std_age_at_death = round(sd(recode_age_int,
                                  na.rm = T), 1),
            .groups = "drop") %>% 
  mutate(`mean (sd)` = paste0(mean_age_at_death, " (", std_age_at_death, ")")) %>% 
  select(census_occ_title, `mean (sd)`) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `mean (sd)`) %>% 
  mutate(measure = "Age at Death",
         level = "Overall") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_age_at_death %>% 
  dt_table()
```

### By Occupation and Sex

```{r}
#| output: true
c1_age_at_death_sex = df_cohort1 %>% 
  group_by(census_occ_title, recode_sex) %>% 
  summarize(mean_age_at_death = round(mean(recode_age_int, 
                                      na.rm = T), 1),
            std_age_at_death = round(sd(recode_age_int,
                                  na.rm = T), 1),
            .groups = "drop") %>% 
  mutate(`mean (sd)` = paste0(mean_age_at_death, " (", std_age_at_death, ")")) %>% 
  select(census_occ_title, recode_sex, `mean (sd)`) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `mean (sd)`) %>% 
  rename(level = recode_sex) %>% 
  mutate(measure = "Age at Death") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_age_at_death_sex %>% 
  dt_table()
```

### Categorical
```{r}
#| output: true
c1_age_at_death_cat = df_cohort1 %>% 
  mutate(age_cat = case_when(recode_age_int < 40 ~ "1. <40",
                             recode_age_int >=40 & recode_age_int <=55 ~ "2. 40-55",
                             recode_age_int > 55 & recode_age_int <=64 ~ "3. 55 - 64",
                             recode_age_int > 64 ~ "4. 65+",
                             is.na(recode_age_int) ~ "5. Unknown",
                             TRUE ~ "ISSUE")) %>% 
  group_by(census_occ_title, age_cat) %>% 
  summarize(num_in_category = n_distinct(id),
            .groups = "drop_last") %>% 
  mutate(`percent in category` = as.character(round(num_in_category / sum(num_in_category) * 100, 2))) %>% 
  select(census_occ_title, age_cat, `percent in category`) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `percent in category`) %>% 
  mutate(measure = "Age at Death Percentage") %>% 
  rename(level = age_cat) %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_age_at_death_cat %>% 
  dt_table()
```

### COVID-19

#### Age at Death by Pre vs Post COVID Overall
```{r}
#| output: true
df_cohort1 %>% 
  mutate(age_covid = case_when(date.of.death <= mdy("02-29-2020") ~ "Pre-Covid",
                             date.of.death > mdy("02-29-2020") ~ "Post-Covid",
                             TRUE ~ "ISSUE")) %>% 
  group_by(age_covid) %>% 
  summarize(mean_age_at_death = round(mean(recode_age_int, 
                                      na.rm = T), 1),
            std_age_at_death = round(sd(recode_age_int,
                                  na.rm = T), 1),
            .groups = "drop") %>% 
  mutate(`mean (sd)` = paste0(mean_age_at_death, " (", std_age_at_death, ")")) %>% 
  select(age_covid,
         `mean (sd)`) %>% 
  dt_table(page = F)
```


#### Age at Death by Pre vs Post COVID by Occupation
```{r}
#| output: true
df_cohort1 %>% 
  mutate(age_covid = case_when(date.of.death <= mdy("02-29-2020") ~ "Pre-Covid",
                             date.of.death > mdy("02-29-2020") ~ "Post-Covid",
                             TRUE ~ "ISSUE")) %>% 
  group_by(census_occ_title, age_covid) %>% 
  summarize(mean_age_at_death = round(mean(recode_age_int, 
                                      na.rm = T), 1),
            std_age_at_death = round(sd(recode_age_int,
                                  na.rm = T), 1),
            .groups = "drop") %>% 
  mutate(`mean (sd)` = paste0(mean_age_at_death, " (", std_age_at_death, ")")) %>% 
  select(census_occ_title,
         age_covid,
         `mean (sd)`) %>% 
  dt_table(page = F)
```

#### Age at Death by Pre vs Post COVID by Occupation and Sex
```{r}
#| output: true
df_cohort1 %>% 
  mutate(age_covid = case_when(date.of.death <= mdy("02-29-2020") ~ "1. Pre-Covid",
                             date.of.death > mdy("02-29-2020") ~ "2. Post-Covid",
                             TRUE ~ "ISSUE")) %>% 
  group_by(census_occ_title, recode_sex, age_covid) %>% 
  summarize(n = n(),
            mean_age_at_death = round(mean(recode_age_int, 
                                      na.rm = T), 1),
            std_age_at_death = round(sd(recode_age_int,
                                  na.rm = T), 1),
            .groups = "drop") %>% 
  mutate(`mean (sd)` = paste0(mean_age_at_death, " (", std_age_at_death, ")")) %>% 
  select(census_occ_title,
         recode_sex,
         age_covid,
         n,
         `mean (sd)`) %>% 
  arrange(census_occ_title, recode_sex, age_covid) %>% 
  dt_table(page = F)
```







#### Percent Deaths Pre vs Post COVID by Occupation Overall
```{r}
#| output: true
c1_percent_covid_overall = df_cohort1 %>% 
  mutate(age_covid = case_when(date.of.death <= mdy("02-29-2020") ~ "Pre-Covid",
                             date.of.death > mdy("02-29-2020") ~ "Post-Covid",
                             TRUE ~ "ISSUE")) %>% 
  group_by(census_occ_title, age_covid) %>% 
  summarize(num_in_category = n_distinct(id),
            .groups = "drop_last") %>% 
  mutate(`percent in category` = as.character(round(num_in_category / sum(num_in_category) * 100, 2))) %>%  
  select(census_occ_title, age_covid, `percent in category`) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `percent in category`) %>% 
  mutate(measure = "Age at Death Percentage") %>% 
  rename(level = age_covid) %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_percent_covid_overall %>% 
  dt_table(page = F)
```

#### Deaths by Year Overall
```{r}
#| output: true
df_cohort1 %>% 
  mutate(year = year(date.of.death)) %>% 
  count(year) %>% 
  arrange(year) %>% 
  dt_table(page = F) %>% 
  add_commas("n")
```

#### Deaths by Year by Occupation
```{r}
#| output: true
deaths_year_occ = df_cohort1 %>% 
  mutate(year = year(date.of.death)) %>% 
  group_by(census_occ_title) %>% 
  count(year) %>% 
  ungroup() %>% 
  pivot_wider(.,
              names_from = "census_occ_title",
              values_from = "n") %>% 
  rename(manager_deaths = `Farmers, Ranchers, and Other Agricultural Managers`,
         worker_deaths = `Miscellaneous Agricultural Workers`) %>% 
  mutate(percent_change_manager = round((manager_deaths - lag(manager_deaths)) / lag(manager_deaths) * 100, 2),
         percent_change_worker = round((worker_deaths - lag(worker_deaths)) / lag(worker_deaths) * 100, 2))
  
deaths_year_occ %>% 
  dt_table(page = F) 
```

#### Plot of Deaths by Year By Occupation
```{r}
#| output: true
deaths_year_occ %>% 
  ggplot() +
  geom_line(aes(x = year,
                y = manager_deaths,
            color = "manager_deaths")) +
  geom_line(aes(x = year,
                y = worker_deaths,
            color = "worker_deaths")) +
  theme_minimal()
```


#### Plot of Percent Change in Deaths by Year By Occupation
```{r}
#| output: true
deaths_year_occ %>% 
  ggplot() +
  geom_line(aes(x = year,
                y = percent_change_manager ,
            color = "manager_deaths_percent_change")) +
  geom_line(aes(x = year,
                y = percent_change_worker,
            color = "worker_deaths_percent_change")) +
  ylab("Percent Change") +
  scale_y_continuous(breaks = seq(-50, 50, 20),) +
  theme_minimal()
```


### Distributions

#### Overall
```{r}
#| output: true
aad_overall_dist = df_cohort1 %>% 
  ggplot(.,
         aes(recode_age_int)) +
  geom_histogram(bins = 101) +
  scale_x_continuous(breaks = seq(0, 120, 15)) +
  ylab("Count of Deaths") +
  xlab("Age at Death") +
  ggtitle("Cohort 1 Age at Death Overall")
  
aad_overall_dist
```

#### By Occupation (Count of Deaths)
```{r}
#| output: true
aad_occupation_count = df_cohort1 %>% 
  ggplot(.,
         aes(recode_age_int)) +
  geom_histogram(bins = 101) +
  scale_x_continuous(breaks = seq(0, 120, 15)) +
  ylab("Count of Deaths") +
  xlab("Age at Death") +
  ggtitle("Cohort 1 Age at Death by Occupation") +
  facet_wrap(~census_occ_title)
  
aad_occupation_count
```


#### By Occupation (Percent of Deaths)
```{r}
#| output: true
aad_occupation_pct = df_cohort1 %>% 
  ggplot(.,
         aes(recode_age_int)) +
  geom_histogram(bins = 101,
                 aes(y = after_stat(count / sum(count) * 100))) +
  scale_x_continuous(breaks = seq(0, 120, 15)) +
  ylab("Percent of Deaths") +
  xlab("Age at Death") +
  ggtitle("Cohort 1 Age at Death by Occupation") +
  facet_wrap(~census_occ_title)
  
aad_occupation_pct
```
  
  
```{r}
#| output: true
aad_occ_violin = df_cohort1 %>% 
  filter(!is.na(recode_age_int)) %>% 
  ggplot(.,
         aes(x = census_occ_title,
             y = recode_age_int)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  scale_y_continuous(breaks = seq(0, 120, 10)) +
  ylab("Age at Death") +
  xlab("Occupation") +
  ggtitle("Cohort 1 Age at Death Overall")
  
aad_occ_violin
```

#### Kolmogorov-Smirnov Test
```{r}
ks_distributions = df_cohort1 %>% 
  group_by(census_occ_title, recode_age_int) %>% 
  summarize(n = n(),
            .groups = "drop") %>% 
  pivot_wider(names_from = "census_occ_title",
              values_from = "n") %>% 
  filter(!is.na(`Farmers, Ranchers, and Other Agricultural Managers`) &
         !is.na(`Miscellaneous Agricultural Workers`))
```

##### Alternative: Two-Sided
```{r}
#| output: true
ks.test(x = ks_distributions$`Farmers, Ranchers, and Other Agricultural Managers`,
        y = ks_distributions$`Miscellaneous Agricultural Workers`,
        alternative = "greater")
```

##### Alternative: Greater
```{r}
#| output: true
ks.test(x = ks_distributions$`Farmers, Ranchers, and Other Agricultural Managers`,
        y = ks_distributions$`Miscellaneous Agricultural Workers`,
        alternative = "greater")
```

#### Cumulative Relative Frequencies

```{r}
crfs = df_cohort1 %>% 
  group_by(census_occ_title, recode_age_int) %>% 
  summarize(n = n(),
            .groups = "drop") %>% 
  pivot_wider(names_from = "census_occ_title",
              values_from = "n") %>% 
  filter(!is.na(`Farmers, Ranchers, and Other Agricultural Managers`) &
         !is.na(`Miscellaneous Agricultural Workers`)) %>% 
  mutate(percent_manager = round(`Farmers, Ranchers, and Other Agricultural Managers` / sum(`Farmers, Ranchers, and Other Agricultural Managers`) * 100, 2),
         cumulative_percent_manager = cumsum(percent_manager),
         percent_worker = round(`Miscellaneous Agricultural Workers` / sum(`Miscellaneous Agricultural Workers`) * 100, 2),
         cumulative_percent_worker = cumsum(percent_worker))

```

```{r}
#| output: true
crfs %>% 
  select(recode_age_int,
         cumulative_percent_worker,
         cumulative_percent_manager) %>% 
  rename(manager = cumulative_percent_manager,
         worker = cumulative_percent_worker) %>% 
  pivot_longer(cols = c("manager", "worker"),
               values_to = "cumulative_percent",
               names_to = "group") %>% 
  ggplot(., aes(x = recode_age_int,
                y = cumulative_percent,
                color = group)) +
  geom_freqpoly(stat = "identity") +
  scale_x_continuous(breaks = seq(1,115, 10)) +
  ggtitle("Cumulative Relative Frequency of Deaths by Occupation") +
  ylab("Percent of Deaths") +
  xlab("Age")
```


##### By Occupation (Percentage)
```{r}
#| output: true
aad_occupation_count_65 = df_cohort1 %>% 
  filter(recode_age_int < 65) %>% 
  ggplot(.,
         aes(recode_age_int)) +
  geom_histogram(bins = 51,
                 aes(y = after_stat(count / sum(count) * 100))) +
  scale_x_continuous(breaks = seq(0, 65, 10)) +
  ylab("Percent of Deaths") +
  xlab("Age at Death") +
  ggtitle("Cohort 1 Age at Death by Occupation") +
  facet_wrap(~census_occ_title)
  
aad_occupation_count_65
```




```{r}
#| output: true
df_cohort1 %>% 
  group_by(census_occ_title, 
           recode_age_int) %>% 
  summarize(n = n(),
            .groups = "drop") %>% 
  ggplot(.,
         aes(x = recode_age_int,
            y = n)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = seq(0, 120, 15)) +
  facet_wrap(~census_occ_title)
```






## Country of Birth

### Overall
```{r}
#| output: true
df_cohort1 %>% 
  mutate(country_of_birth = case_when(birth.country == "US" ~ "1. United States",
                                      birth.country == "MX" ~ "2. Mexico",
                                      TRUE ~ "3.Other")) %>% 
  count(country_of_birth) %>% 
  dt_table() %>% 
  add_commas("n")
```

### By Occupation
```{r}
#| output: true
c1_country_birth = df_cohort1 %>% 
  mutate(country_of_birth = case_when(birth.country == "US" ~ "1. United States",
                                      birth.country == "MX" ~ "2. Mexico",
                                      TRUE ~ "3. Other")) %>% 
  group_by(census_occ_title, country_of_birth) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop_last") %>%
  mutate(percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                          " (",
                                                          percent_within_occupation,
                                                          "%)")) %>%
  ungroup() %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  rename(level = country_of_birth) %>% 
  mutate(measure = "Country of Birth") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_country_birth %>% 
  dt_table(page = F) 
```


## Facility at Death

### Overall
```{r}
#| output: true
df_cohort1 %>% 
  count(recode_facility_of_death) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  arrange(desc(n)) %>% 
  dt_table() %>% 
  add_commas("n")
```


### By Occupation
```{r}
#| output: true
c1_facility_death = df_cohort1 %>% 
  group_by(census_occ_title, recode_facility_of_death) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop_last") %>%
  mutate(percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                          " (",
                                                          percent_within_occupation,
                                                          "%)")) %>%
  ungroup() %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  rename(level = recode_facility_of_death) %>% 
  mutate(measure = "Facility of Death") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_facility_death %>% 
  dt_table(page = F) 
```












## Table 1
```{r}
#| output: true
bind_rows(c1_num_deaths,
          c1_education,
          c1_marital_status,
          c1_sex,
          c1_race,
          c1_age_at_death,
          c1_age_at_death_sex,
          c1_age_at_death_cat,
          c1_country_birth,
          c1_facility_death) %>% 
  dt_table()
```


## Cause of Death

:::{.callout-note}
I manually added some code descriptions to the df_icd table because they did not match the latest tuva ICD description table. These are probably slightly older codes. I also split out the first letter (`start_letter`) as its own variable and the 2nd and 3rd characters of the ICD-10s into a second extra variable (`start_number`) primarily for coding into external causes. I coerced the `start_number` to an integer, which introduced some NAs, but these do not exist in any of the external cause codes (V00 - Y99). See below for start letters that contain null start_numbers.
:::

```{r}
df_cod = df_cohort1 %>% 
  mutate(final.cause.of.death.icd10 = substr(final.cause.of.death.icd10, 1, 3),
         final.cause.of.death.icd10 = if_else(final.cause.of.death.icd10 == "",
                                              NA,
                                              final.cause.of.death.icd10))
```

```{r}
df_icd = read_csv(here::here("data/icd_10_cm_codes_tuva.csv.gz"),
                  col_names = c("code",
                                "billable",
                                "short_desc",
                                "long_desc")) %>% 
         bind_rows(tribble(~code, ~ billable, ~short_desc, ~long_desc,
                   "X70", 0, "Intentional self-harm by hanging, strangulation and suffocation", "",
                   "X44", 0, "Accidental poisoning by and exposure to other and unspecified drugs, medicaments and biological substances", "",
                   "X41", 0, "Accidental poisoning by and exposure to antiepileptic, sedative-hypnotic, antiparkinsonism and psychotropic drugs, not elsewhere classified", "",
                   "X42", 0, "Accidental poisoning by and exposure to narcotics and psychodysleptics [hallucinogens], not elsewhere classified", "")) %>% 
        mutate(start_letter = str_sub(code, 1, 1),
               start_number = as.integer(str_sub(code, 2, 3)),
               external_natural = if_else(start_letter %in% c("V", "W", "X", "Y"),
                                          "External",
                                          "Natural"))
```

Start letters that contain null start_numbers:
```{r}
#| output: true
df_icd %>% 
  filter(is.na(start_number)) %>% 
  distinct(start_letter) %>% 
  arrange(start_letter)
```

```{r}
#| eval: false
# check that no deaths are joining to multiple codes. This should return empty set
df_cod %>% 
  left_join(df_icd,
            by = c("final.cause.of.death.icd10" = "code"),
            keep = TRUE) %>% 
  group_by(id) %>% 
  count() %>% 
  filter(n > 1)
```

### Final Cause of Death

:::{.callout-note}
A number of reported ICD 10 CM codes in the data do not match to a real ICD 10 CM code.
For now, I am shortening the reported codes to 3 characters and then joining to increase matching. Further some final.cause.of.death.icd10 values are empty strings. I'm converting to NA and filtering. 
:::

Number of records with Null cause of death
```{r}
#| output: true
df_cod %>% 
  filter(is.na(final.cause.of.death.icd10)) %>% 
  count()
```

```{r}
# filtering null causes of death
df_cod = df_cod %>% 
  filter(!is.na(final.cause.of.death.icd10))
```

#### Overall
```{r}
#| output: TRUE
df_cod %>%  
  group_by(final.cause.of.death.icd10) %>% 
  summarize(num_deaths = n()) %>% 
  arrange(desc(num_deaths)) %>% 
  left_join(df_icd,
            by = c("final.cause.of.death.icd10" = "code"),
            keep = TRUE) %>% 
  select(final.cause.of.death.icd10,
         code,
         short_desc,
         num_deaths) %>% 
  dt_table()
```


#### Managers
```{r}
#| output: TRUE
df_cod %>%  
  filter(census_occ_title == "Farmers, Ranchers, and Other Agricultural Managers") %>% 
  group_by(final.cause.of.death.icd10) %>% 
  summarize(num_deaths = n()) %>% 
  arrange(desc(num_deaths)) %>% 
  left_join(df_icd,
            by = c("final.cause.of.death.icd10" = "code"),
            keep = TRUE) %>% 
  select(final.cause.of.death.icd10,
         code,
         short_desc,
         num_deaths) %>% 
  dt_table()
```

#### Workers
```{r}
#| output: TRUE
df_cod %>% 
  filter(census_occ_title == "Miscellaneous Agricultural Workers") %>% 
  group_by(final.cause.of.death.icd10) %>% 
  summarize(num_deaths = n()) %>% 
  arrange(desc(num_deaths)) %>% 
  left_join(df_icd,
            by = c("final.cause.of.death.icd10" = "code"),
            keep = TRUE) %>% 
  select(final.cause.of.death.icd10,
         code,
         short_desc,
         num_deaths) %>% 
  dt_table()
```


#### External Vs. Natural Causes

I'm filtering out final causes of death that are NA.

```{r}
#| output: TRUE
df_cod %>% 
  left_join(df_icd,
            by = c("final.cause.of.death.icd10" = "code"),
            keep = TRUE) %>% 
  group_by(census_occ_title, external_natural) %>% 
  count() %>% 
  ungroup() %>% 
  pivot_wider(names_from = "census_occ_title",
              values_from = "n") %>% 
  mutate(manager_percent = round(`Farmers, Ranchers, and Other Agricultural Managers` / sum(`Farmers, Ranchers, and Other Agricultural Managers`) * 100, 2),
         worker_percent = round(`Miscellaneous Agricultural Workers` / sum(`Miscellaneous Agricultural Workers`) * 100, 2)) %>% 
  dt_table()
```


#### Age Under 60

:::{.callout-note}
In the overall cohort, the sample is about 42% managers and 58% workers, but in the under 60 sample, this turns into about 22% managers and 78% workers. The overall sample is 52,786 deaths and the under 60 sample is 8,941 deaths.
:::


##### Percent of Total Cohort
```{r}
#| output: true
df_cohort1 %>% 
  mutate(under_over_60 = if_else(recode_age_int < 60, "Under 60", "60 or Over")) %>%
  count(under_over_60) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  dt_table()
```

##### Percent of Total Cohort by Occupation
```{r}
#| output: true
df_cohort1 %>% 
  mutate(under_over_60 = if_else(recode_age_int < 60, "Under 60", "60 or Over")) %>%
  count(census_occ_title, under_over_60) %>% 
  pivot_wider(names_from = "census_occ_title",
              values_from = "n") %>% 
  mutate(manager_percent = round(`Farmers, Ranchers, and Other Agricultural Managers` / sum(`Farmers, Ranchers, and Other Agricultural Managers`) * 100, 2),
         worker_percent = round(`Miscellaneous Agricultural Workers` / sum(`Miscellaneous Agricultural Workers`) * 100, 2)) %>% 
  dt_table()
```



##### Sample Sizes
```{r}
#| output: true
df_cod %>% 
  filter(recode_age_int < 60) %>% 
  group_by(census_occ_title) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop") %>% 
  mutate(measure = "Number of Deaths",
         level = "Overall",
         percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                         " (",
                                                         percent_within_occupation,
                                                         "%)")) %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  dt_table()
```

##### Overall
```{r}
#| output: TRUE
df_cod %>% 
  filter(recode_age_int < 60) %>%  
  group_by(final.cause.of.death.icd10) %>% 
  summarize(num_deaths = n()) %>% 
  arrange(desc(num_deaths)) %>% 
  left_join(df_icd,
            by = c("final.cause.of.death.icd10" = "code"),
            keep = TRUE) %>% 
  select(final.cause.of.death.icd10,
         code,
         short_desc,
         num_deaths) %>% 
  dt_table()
```

##### Managers
```{r}
#| output: TRUE
df_cod %>% 
  filter(recode_age_int < 60) %>% 
  filter(census_occ_title == "Farmers, Ranchers, and Other Agricultural Managers") %>% 
  group_by(final.cause.of.death.icd10) %>% 
  summarize(num_deaths = n()) %>% 
  arrange(desc(num_deaths)) %>% 
  left_join(df_icd,
            by = c("final.cause.of.death.icd10" = "code"),
            keep = TRUE) %>% 
  select(final.cause.of.death.icd10,
         code,
         short_desc,
         num_deaths) %>% 
  dt_table()
```

##### Workers
```{r}
#| output: TRUE
df_cod %>% 
  filter(recode_age_int < 60) %>%
  filter(census_occ_title == "Miscellaneous Agricultural Workers") %>% 
  group_by(final.cause.of.death.icd10) %>% 
  summarize(num_deaths = n()) %>% 
  arrange(desc(num_deaths)) %>% 
  left_join(df_icd,
            by = c("final.cause.of.death.icd10" = "code"),
            keep = TRUE) %>% 
  select(final.cause.of.death.icd10,
         code,
         short_desc,
         num_deaths) %>% 
  dt_table()
```

##### External Vs. Natural Causes
```{r}
#| output: TRUE
df_cod %>% 
  filter(recode_age_int < 60) %>%
  left_join(df_icd,
            by = c("final.cause.of.death.icd10" = "code"),
            keep = TRUE) %>% 
  group_by(census_occ_title, external_natural) %>% 
  count() %>% 
  ungroup() %>% 
  pivot_wider(names_from = "census_occ_title",
              values_from = "n") %>% 
  mutate(manager_percent = round(`Farmers, Ranchers, and Other Agricultural Managers` / sum(`Farmers, Ranchers, and Other Agricultural Managers`) * 100, 2),
         worker_percent = round(`Miscellaneous Agricultural Workers` / sum(`Miscellaneous Agricultural Workers`) * 100, 2)) %>% 
  dt_table()
```












