---
title: "Class and Death Project Development Notebook"
subtitle: "Exploration of cohort creation prior to analysis"
author: "Luke Slipski"
date: "Published on `r format(Sys.time(), '%m-%d-%y')`"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 5
fig-cap-location: top
editor:
  markdown:
    wrap: 72
execute: 
  echo: false
  output: false
---

```{r}
library("tidyverse")
# download CAdeathserver with remotes::install_git("https://github.com/Lslipski/CAdeathserver")
library("CAdeathserver")
# download merced.clc package with remotes::install_git("https://github.com/Lslipski/merced.clc")
library("merced.clc")
```

# Main Data Loading

The main data set includes the following filters:

  -   State of residence is CA
  
```{r}
# Get clean mortality dataset
# This file is large and will take a minute to load
df_mortality_clean = load_ca_mortality_data(df = 'latest') %>% 
  get_ca_residence_only() %>% 
  recode_race() %>%
  recode_age() %>%
  recode_education() %>%
  recode_sex() %>% 
  create_covid_death_flag()
```


# Industry and Occupation Code Cleaning

I used the below code to remove industry and occupation fields that contained the `&` and `#` characters
because there were duplicates with these when using `load_industry_occupation()`. This updated data file was saved to `E:/shared/heb_lab/data/nioccs/nioccs_coded_ind_occ_20230831_mortality.rds`, which is currently where
the CAdeathserver package loads industry and occupation data from.

```{r}
#| echo: true
#| eval: false
target <- c("&", "#") %>% paste(collapse = "|")

ind_occ_to_remove = load_industry_occupation() %>% 
  filter(if_any(contains("ind"), 
                ~str_detect(.x, target)) |
           if_any(contains("occ"), 
                ~str_detect(.x, target)))

df_ind_occ_all %>% saveRDS(.,
                           file = "E:/shared/heb_lab/data/nioccs/nioccs_coded_ind_occ_20230831_mortality.rds")
```

# Group Variables

## Class 

This section takes the data set that was returned from the NIOCCS tool, which
has a record for every unique industry + occupation combination in the original
mortality data from 2023-08-31, and looks at the various industry and occupation
standards to which the NIOCCS tool mapped the original mortality combinations. These
are SOC (standard occupation classification) and Census codes. 

Main Findings: 
  -   SOC and census occupation are identically missing (~21k records)
  -   There are fewer records missing census industry than census occupation, it is still too many to manually re-code (~15k records) 
  -   The number of records missing BOTH census industry and census occupation information is much lower (1,650)

## Industry Codes

```{r}
df_indocc <- df_mortality_clean %>% 
  join_industry_occupation(.) %>% 
  select(id = id.x,
         industry,
         occupation,
         starts_with("census"),
         starts_with("naics"),
         starts_with("soc_"),
         unexpected_naics_soc_combo) %>% 
  distinct()
```















