---
title: "Class and Death Project Development Notebook"
subtitle: "Exploration of cohort creation prior to analysis"
author: "Luke Slipski"
date: "Published on `r format(Sys.time(), '%m-%d-%y')`"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 5
fig-cap-location: top
editor:
  markdown:
    wrap: 72
execute: 
  echo: false
  output: false
---

```{r}
library("tidyverse")
# download CAdeathserver with remotes::install_git("https://github.com/Lslipski/CAdeathserver")
library("CAdeathserver")
# download merced.clc package with remotes::install_git("https://github.com/Lslipski/merced.clc")
library("merced.clc")
```

# Main Data Loading

The main data set includes the following filters:

  -   State of residence is CA
  
```{r}
# Get clean mortality dataset
# This file is large and will take a minute to load
df_mortality_clean = load_ca_mortality_data(df = 'latest') %>% 
  get_ca_residence_only() %>% 
  recode_race() %>%
  recode_age() %>%
  recode_education_farmworkers() %>%
  recode_sex() %>% 
  create_covid_death_flag() %>% 
  recode_marital_status() %>% 
  join_industry_occupation()
```


# Industry and Occupation Code Cleaning

I used the below code to remove industry and occupation fields that contained the `&` and `#` characters
because there were duplicates with these when using `load_industry_occupation()`. This updated data file was saved to `E:/shared/heb_lab/data/nioccs/nioccs_coded_ind_occ_20230831_mortality.rds`, which is currently where
the CAdeathserver package loads industry and occupation data from.

```{r}
#| echo: true
#| eval: false
# target <- c("&", "#") %>% paste(collapse = "|")
# 
# ind_occ_to_remove = load_industry_occupation() %>% 
#   filter(if_any(contains("ind"), 
#                 ~str_detect(.x, target)) |
#            if_any(contains("occ"), 
#                 ~str_detect(.x, target)))
# 
# df_ind_occ_all %>% saveRDS(.,
#                            file = "E:/shared/heb_lab/data/nioccs/nioccs_coded_ind_occ_20230831_mortality.rds")
```

# Group Variables

## Industry and Occupation 

```{r}
df_indocc <- df_mortality_clean %>% 
  select(id,
         industry,
         occupation,
         starts_with("census"),
         starts_with("naics"),
         starts_with("soc_"),
         unexpected_naics_soc_combo) %>% 
  distinct()
```

### Summary Review of Code Missingness

This section takes the data set that was returned from the NIOCCS tool, which
has a record for every unique industry + occupation combination in the original
mortality data from 2023-08-31, and looks at the various industry and occupation
standards to which the NIOCCS tool mapped the original mortality combinations. These
are SOC (standard occupation classification) and Census codes. 

Main Findings: 
  -   SOC and census occupation are identically missing (~21k records)
  -   There are fewer records missing census industry than census occupation, it is still too many to manually re-code (~15k records) 
  -   The number of records missing BOTH census industry and census occupation information is much lower (1,650)


### Codes of Interest

Descriptions taken from [here](https://www2.census.gov/programs-surveys/cps/methodology/Occupation%20Codes.pdf)

```{r}
#| output: true
occ_codes_of_interest = tribble(~census_occ_code_2018, ~description,
                                "6050", "Miscellaneous Agricultural Workers",
                                "6040", "Graders and sorters, agricultural products",
                                "6005", "First-line supervisors of farming, fishing, and forestry workers",
                                "6010", "Agricultural inspectors",
                                "6020", "Animal breeders",
                                "6115", "Fishing and hunting workers",
                                "6120", "Forest and conservation workers",
                                "6130", "Logging workers",
                                "0205", "Farmers, ranchers, and other agricultural managers",
                                "7810", "Butchers and other meat, poultry, and fish processing workers")

occ_codes_of_interest %>% 
  arrange(census_occ_code_2018) %>% 
  dt_table(search = F,
           page = F)
```


### Count of Mortality Records by Codes of Interest
```{r}
#| output: true
df_indocc %>% 
  filter(census_occ_code %in% occ_codes_of_interest$census_occ_code_2018) %>% 
  count(census_occ_code, 
        census_occ_title) %>% 
  arrange(desc(n)) %>% 
  dt_table() %>% 
  add_commas("n")
```


## Race


### Overall Cohort
```{r}
#| output: true
df_mortality_clean %>% 
  count(recode_race) %>% 
  arrange(desc(n)) %>% 
  dt_table(search = F,
           page = F) %>% 
  add_commas("n")
```

### Limited to Occupations of Interest
```{r}
#| output: true
df_mortality_clean %>% 
  filter(census_occ_code %in% occ_codes_of_interest$census_occ_code_2018) %>% 
  count(census_occ_code,
        census_occ_title,
        recode_race) %>% 
  arrange(census_occ_code,
          census_occ_title,
          desc(n)) %>% 
  dt_table(search = F) %>% 
  add_commas("n")
```



# Cohort 1: Limited to Ag Workers and Ag Managers

This section will limit the cohort to only 2 occupation codes: 
```{r}
#| output: true
cohort1_occ_codes = c("6050", "0205")
occ_codes_of_interest %>% 
  filter(census_occ_code_2018 %in% cohort1_occ_codes) %>% 
  dt_table(search = F,
           page = F)
```

```{r}
df_cohort1 = df_mortality_clean %>% 
  filter(census_occ_code %in% cohort1_occ_codes)
```

## Number of Deaths
```{r}
#| output: true
c1_num_deaths = df_cohort1 %>% 
  group_by(census_occ_title) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop") %>% 
  mutate(measure = "Number of Deaths",
         level = "Overall",
         percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                         " (",
                                                         percent_within_occupation,
                                                         "%)")) %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`)

c1_num_deaths %>% 
  dt_table() 
  
```



## Education

### Overall
```{r}
#| output: true
df_cohort1 %>% 
  count(recode_educ_farmwork) %>% 
  dt_table() %>% 
  add_commas("n")
```
### By Occupation
```{r}
#| output: true
c1_education = df_cohort1 %>% 
  group_by(census_occ_title, recode_educ_farmwork) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop_last") %>%
  mutate(percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                          " (",
                                                          percent_within_occupation,
                                                          "%)")) %>%
  ungroup() %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  rename(level = recode_educ_farmwork) %>% 
  mutate(measure = "Highest Level of Education") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_education %>% 
  dt_table(page = F) 
```












### By Occupation
```{r}
#| output: true
c1_education_age = df_cohort1 %>% 
  filter(recode_age_int >= 25) %>% 
  group_by(census_occ_title, recode_educ_farmwork) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop_last") %>%
  mutate(percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                          " (",
                                                          percent_within_occupation,
                                                          "%)")) %>%
  ungroup() %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  rename(level = recode_educ_farmwork) %>% 
  mutate(measure = "Highest Level of Education") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_education_age %>% 
  dt_table(page = F) 
```














## Marital Status

:::{.callout-note}
We may want to combine married/widowed/divorced with the domestic partnership versions of these.
:::

### Overall
```{r}
#| output: true
df_cohort1 %>% 
  count(recode_marital_status) %>% 
  dt_table() %>% 
  add_commas("n")
```


### By Occupation
```{r}
#| output: true
c1_marital_status = df_cohort1 %>% 
  group_by(census_occ_title, recode_marital_status) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop_last") %>%
  mutate(percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                          " (",
                                                          percent_within_occupation,
                                                          "%)")) %>%
  ungroup() %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  rename(level = recode_marital_status) %>% 
  mutate(measure = "Marital Status") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_marital_status %>% 
  dt_table(page = F) 
```

## Sex

### Overall
```{r}
#| output: true
df_cohort1 %>% 
  count(recode_sex) %>% 
  dt_table() %>% 
  add_commas("n")
```

### By Occupation
```{r}
#| output: true
c1_sex = df_cohort1 %>% 
  group_by(census_occ_title, recode_sex) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop_last") %>%
  mutate(percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                          " (",
                                                          percent_within_occupation,
                                                          "%)")) %>%
  ungroup() %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  rename(level = recode_sex) %>% 
  mutate(measure = "Sex") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_sex %>% 
  dt_table(page = F) 
```

## Race

:::{.callout-note}
We have a number of race codes that we still need to categorize. They're spelled out in the Class and Death Shared Notebook google doc
:::

### Overall
```{r}
#| output: true
df_cohort1 %>% 
  count(recode_race) %>% 
  dt_table() %>% 
  add_commas("n")
```


### By Occupation
```{r}
#| output: true
c1_race = df_cohort1 %>% 
  group_by(census_occ_title, recode_race) %>% 
  summarize(n = n_distinct(id),
            .groups = "drop_last") %>%
  mutate(percent_within_occupation = round(n / sum(n) * 100, 1),
         `Number of workers (Percent within occ)` = paste0(format(n, big.mark = ","), 
                                                          " (",
                                                          percent_within_occupation,
                                                          "%)")) %>%
  ungroup() %>% 
  select(-n, -percent_within_occupation) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `Number of workers (Percent within occ)`) %>% 
  rename(level = recode_race) %>% 
  mutate(measure = "Race") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_race %>% 
  dt_table(page = F) 
```





## Age at Death


### Overall
```{r}
#| output: true
df_cohort1 %>% 
  summarize(mean_age_at_death = round(mean(recode_age_int, 
                                      na.rm = T), 1),
            std_age_at_death = round(sd(recode_age_int,
                                  na.rm = T), 1),
            .groups = "drop") %>% 
  mutate(`mean (sd)` = paste0(mean_age_at_death, " (", std_age_at_death, ")")) %>% 
  select(`mean (sd)`)
```


### By Occupation

```{r}
#| output: true
c1_age_at_death = df_cohort1 %>% 
  group_by(census_occ_title) %>% 
  summarize(mean_age_at_death = round(mean(recode_age_int, 
                                      na.rm = T), 1),
            std_age_at_death = round(sd(recode_age_int,
                                  na.rm = T), 1),
            .groups = "drop") %>% 
  mutate(`mean (sd)` = paste0(mean_age_at_death, " (", std_age_at_death, ")")) %>% 
  select(census_occ_title, `mean (sd)`) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `mean (sd)`) %>% 
  mutate(measure = "Age at Death",
         level = "Overall") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_age_at_death %>% 
  dt_table()
```

### By Occupation and Sex

```{r}
#| output: true
c1_age_at_death_sex = df_cohort1 %>% 
  group_by(census_occ_title, recode_sex) %>% 
  summarize(mean_age_at_death = round(mean(recode_age_int, 
                                      na.rm = T), 1),
            std_age_at_death = round(sd(recode_age_int,
                                  na.rm = T), 1),
            .groups = "drop") %>% 
  mutate(`mean (sd)` = paste0(mean_age_at_death, " (", std_age_at_death, ")")) %>% 
  select(census_occ_title, recode_sex, `mean (sd)`) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `mean (sd)`) %>% 
  rename(level = recode_sex) %>% 
  mutate(measure = "Age at Death") %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_age_at_death_sex %>% 
  dt_table()
```

### Categorical
```{r}
#| output: true
c1_age_at_death_cat = df_cohort1 %>% 
  mutate(age_cat = case_when(recode_age_int < 40 ~ "1. <40",
                             recode_age_int >=40 & recode_age_int <=55 ~ "2. 40-55",
                             recode_age_int > 55 & recode_age_int <=64 ~ "3. 55 - 64",
                             recode_age_int > 64 ~ "4. 65+",
                             is.na(recode_age_int) ~ "5. Unknown",
                             TRUE ~ "ISSUE")) %>% 
  group_by(census_occ_title, age_cat) %>% 
  summarize(num_in_category = n_distinct(id),
            .groups = "drop_last") %>% 
  mutate(`percent in category` = as.character(round(num_in_category / sum(num_in_category) * 100, 2))) %>% 
  select(census_occ_title, age_cat, `percent in category`) %>% 
  pivot_wider(names_from = census_occ_title,
              values_from = `percent in category`) %>% 
  mutate(measure = "Age at Death Percentage") %>% 
  rename(level = age_cat) %>% 
  relocate(measure, level) %>% 
  arrange(level)
  
c1_age_at_death_cat %>% 
  dt_table()
```


## Table 1
```{r}
#| output: true
bind_rows(c1_num_deaths,
          c1_education,
          c1_education_age,
          c1_marital_status,
          c1_sex,
          c1_race,
          c1_age_at_death,
          c1_age_at_death_sex,
          c1_age_at_death_cat) %>% 
  dt_table()
```








