---
title: "Class and Death: Indirect Standardization Notebook"
author: "Luke Slipski"
date: "Published on `r format(Sys.time(), '%m-%d-%y')`"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 5
    page-layout: full
fig-cap-location: top
editor:
  markdown:
    wrap: 72
execute: 
  echo: false
  output: false
---

Following the procedure here for indirect standardization: https://pmc.ncbi.nlm.nih.gov/articles/PMC3406211/

```{r}
library("tidyverse")
library("ipumsr")
library("lubridate")
library("scales")
# download CAdeathserver with remotes::install_git("https://github.com/Lslipski/CAdeathserver")
library("CAdeathserver")
# download merced.clc package with remotes::install_git("https://github.com/Lslipski/merced.clc")
library("merced.clc")
# for working with ICD codes
# install with 
# remotes::install_local("C:\\Users\\luke_slipski\\Downloads\\icd-main.zip")
# after downloading to local from https://github.com/cran/icd
library("icd") 
library("epitools")
```
# Load
```{r}
df_cohort1 = read_rds(here::here("data/df_cohort1.rds")) %>% 
  mutate(month_of_death = as.integer(month.of.death),
         worker_manager = if_else(census_occ_code == "6050",
                                  "Worker",
                                  "Manager"),
         white_non_white = if_else(recode_race == "White", "White", "Non-White")) 

df_ipums = read_rds(here::here("data/demographics.rds")) %>% 
  # remove ipums labels which interact oddly with dt_table
  mutate(dplyr::across(.cols = dplyr::everything(),
                       .fns = unclass),
         sex = if_else(sex == 1 , "Male", "Female"),
         worker_manager = if_else(occ == 6050, "Worker", "Manager"),
         age_cat = case_when(age < 40 ~ "1. <40",
                             age >=40 & age <=55 ~ "2. 40-55",
                             age > 55 & age <=64 ~ "3. 56-64",
                             age > 64 ~ "4. 65+",
                             is.na(age) ~ "5. Unknown",
                             TRUE ~ "ISSUE"),
         hispan = case_when(hispan == 0 ~ "Not Hispanic",
                            hispan == 1 ~ "Mexican",
                            hispan == 2 ~ "Puerto Rican",
                            hispan == 3 ~ "Cuban",
                            hispan == 4 ~ "Other",
                            hispan == 9 ~ "Not reported",
                            TRUE ~ "ISSUE"),
         race = case_when(hispan %in% c(1,2,3,4) ~ "Hispanic",
                          race == 1 ~ "White",
                          race == 2 ~ "Black",
                          race == 3 ~ "American Indian / Native American",
                          race %in% c(4,5,6) ~ "Asian / Pacific Islander",
                          TRUE ~ "Other"),
         white_non_white = if_else(race == "White", "White", "Non-White"))

# limited cohorts restricted to age 25-64 and ipums limited to in the labor force
df_cohort1_restricted = df_cohort1 %>% 
  filter(recode_age_int >= 25 &
         recode_age_int <= 64) %>% 
  mutate(age_cat_every_5 = case_when(
                             recode_age_int < 30 ~ "1. 25 - 29",
                             recode_age_int <= 34 ~ "2. 30 - 34",
                             recode_age_int <= 39 ~ "3. 35 - 39",
                             recode_age_int <= 44 ~ "4. 40 - 44",
                             recode_age_int <= 49 ~ "5. 45 - 49",
                             recode_age_int <= 54 ~ "6. 50 - 54",
                             recode_age_int <= 59 ~ "7. 55 - 59",
                             recode_age_int <= 64 ~ "8. 60 - 64",
                             is.na(recode_age_int) ~ "9. Unknown",
                             TRUE ~ "ISSUE"))
  
df_ipums_restricted = df_ipums %>% 
  filter(age >= 25 &
         age <= 64) %>% 
  mutate(age_cat_every_5 = case_when(
                             age < 30 ~ "1. 25 - 29",
                             age <= 34 ~ "2. 30 - 34",
                             age <= 39 ~ "3. 35 - 39",
                             age <= 44 ~ "4. 40 - 44",
                             age <= 49 ~ "5. 45 - 49",
                             age <= 54 ~ "6. 50 - 54",
                             age <= 59 ~ "7. 55 - 59",
                             age <= 64 ~ "8. 60 - 64",
                             is.na(age) ~ "9. Unknown",
                             TRUE ~ "ISSUE"))

# employment status limit LABFORCE == 2
df_ipums_restricted_emp = df_ipums %>% 
  filter(age >= 25 &
         age <= 64 &
         labforce == 2) %>% 
  mutate(age_cat_every_5 = case_when(
                             age < 30 ~ "1. 25 - 29",
                             age <= 34 ~ "2. 30 - 34",
                             age <= 39 ~ "3. 35 - 39",
                             age <= 44 ~ "4. 40 - 44",
                             age <= 49 ~ "5. 45 - 49",
                             age <= 54 ~ "6. 50 - 54",
                             age <= 59 ~ "7. 55 - 59",
                             age <= 64 ~ "8. 60 - 64",
                             is.na(age) ~ "9. Unknown",
                             TRUE ~ "ISSUE"))
```

## Indirect Standardization

### Age -- Workers (Reference) Vs. Managers 
```{r}
# group deaths by year, worker/manager, and age category and count the number of 
# deaths in each strata
df_deaths_per_stratum = df_cohort1_restricted %>%
  group_by(year.of.death,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_deaths = n_distinct(id),
            .groups = "drop") %>% 
  rename(year = year.of.death)
  
# group population by year, worker/manager, and age category and count the number
# of people in each strata
df_people_per_stratum = df_ipums_restricted_emp %>% 
  group_by(year,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_people = sum(perwt),
            .groups = "drop")
  
# get worker death rates by year and age group by joining to the number of people
# in worker strata and dividing deaths by people
df_worker_death_rates_crude = df_deaths_per_stratum %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  mutate(death_rate_crude_worker = sum_deaths / sum_people)

# calculate crude worker death rate overall by year for workers
df_worker_rates_crude_year = df_deaths_per_stratum %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  group_by(year) %>% 
  summarize(total_worker_deaths = sum(sum_deaths),
            total_workers = sum(sum_people),
            .groups = "drop") %>% 
  mutate(worker_crude_death_rate = total_worker_deaths / total_workers)
  
# start with number of people per stratum for managers only
# join to get the worker death rates per stratum
# calculate expected deaths as number of managers times worker death rates per stratum
# join to get actual number of manager deaths per stratum
df_manager_rates = df_people_per_stratum %>% 
  filter(worker_manager == "Manager") %>% 
  left_join(df_worker_death_rates_crude %>% 
              select(year, 
                     age_cat_every_5, 
                     death_rate_crude_worker),
            by = c("year", "age_cat_every_5")) %>% 
  mutate(expected_deaths = round(sum_people * death_rate_crude_worker, 2)) %>% 
  left_join(df_deaths_per_stratum %>% 
              filter(worker_manager == "Manager"),
            by = c("year", "worker_manager", "age_cat_every_5"))
  
# start with the manager expected and actual death counts by stratum 
# group into years and sum the expected and actual death counts by year
# divide actual by expected to get the adjusted mortality rate by year
# join back to worker crude rates by year 
df_manager_rates_adj_year = df_manager_rates %>% 
  group_by(year) %>% 
  summarize(total_expected = sum(expected_deaths),
            total_observed = sum(sum_deaths),
            .groups = "drop") %>% 
  mutate(manager_std_mortality_ratio = round(total_observed / total_expected, 2)) %>% 
  left_join(df_worker_rates_crude_year,
           by = "year") %>% 
  mutate(manager_adj_mortality_rate = manager_std_mortality_ratio * worker_crude_death_rate)
```


```{r}
#| output: true
df_manager_rates_adj_year %>% 
  mutate(worker_crude_death_rate = round(worker_crude_death_rate, 5)) %>% 
  relocate(year,
           total_worker_deaths,
           total_workers, 
           worker_crude_death_rate,
           total_expected,
           total_observed,
           manager_std_mortality_ratio,
           manager_adj_mortality_rate) %>% 
  dt_table() %>% 
  add_commas("total_workers")
```


### Age -- Non-White Workers (Reference) vs. Non-White Managers
```{r}
# group deaths by year, worker/manager, and age category
# filter to non-white people only 
# count the number of deaths in each strata
df_deaths_per_stratum = df_cohort1_restricted %>%
  filter(white_non_white == "Non-White") %>% 
  group_by(year.of.death,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_deaths = n_distinct(id),
            .groups = "drop") %>% 
  rename(year = year.of.death)
  
# group population by year, worker/manager, and age category
# filter to non-white people only 
# count the number of people in each strata
df_people_per_stratum = df_ipums_restricted_emp %>% 
  filter(white_non_white == "Non-White") %>% 
  group_by(year,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_people = sum(perwt),
            .groups = "drop")
  
# get worker death rates by year and age group by joining to the number of people
# in worker strata and dividing deaths by people
df_worker_death_rates_crude = df_deaths_per_stratum %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  mutate(death_rate_crude_worker = sum_deaths / sum_people)

# calculate crude worker death rate overall by year for workers
df_worker_rates_crude_year = df_deaths_per_stratum %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  group_by(year) %>% 
  summarize(total_worker_deaths = sum(sum_deaths),
            total_workers = sum(sum_people),
            .groups = "drop") %>% 
  mutate(worker_crude_death_rate = total_worker_deaths / total_workers)
  
# start with number of people per stratum for managers only
# join to get the worker death rates per stratum
# calculate expected deaths as number of managers times worker death rates per stratum
# join to get actual number of manager deaths per stratum
df_manager_rates = df_people_per_stratum %>% 
  filter(worker_manager == "Manager") %>% 
  left_join(df_worker_death_rates_crude %>% 
              select(year, 
                     age_cat_every_5, 
                     death_rate_crude_worker),
            by = c("year", "age_cat_every_5")) %>% 
  mutate(expected_deaths = round(sum_people * death_rate_crude_worker, 2)) %>% 
  left_join(df_deaths_per_stratum %>% 
              filter(worker_manager == "Manager"),
            by = c("year", "worker_manager", "age_cat_every_5"))
  
# start with the manager expected and actual death counts by stratum 
# group into years and sum the expected and actual death counts by year
# divide actual by expected to get the adjusted mortality rate by year
# join back to worker crude rates by year 
df_manager_rates_adj_year = df_manager_rates %>% 
  group_by(year) %>% 
  summarize(total_expected = sum(expected_deaths),
            total_observed = sum(sum_deaths),
            .groups = "drop") %>% 
  mutate(manager_std_mortality_ratio = round(total_observed / total_expected, 2)) %>% 
  left_join(df_worker_rates_crude_year,
           by = "year") %>% 
  mutate(manager_adj_mortality_rate = manager_std_mortality_ratio * worker_crude_death_rate)
```

```{r}
#| output: true
df_manager_rates_adj_year %>% 
  mutate(worker_crude_death_rate = round(worker_crude_death_rate, 5),
         total_expected = round(total_expected,2)) %>% 
  relocate(year,
           total_worker_deaths,
           total_workers, 
           worker_crude_death_rate,
           total_expected,
           total_observed,
           manager_std_mortality_ratio,
           manager_adj_mortality_rate) %>% 
  dt_table() %>% 
  add_commas("total_workers")
```

### Age -- White Workers (Reference) vs. White Managers
```{r}
# group deaths by year, worker/manager, and age category
# filter to non-white people only 
# count the number of deaths in each strata
df_deaths_per_stratum = df_cohort1_restricted %>%
  filter(white_non_white == "White") %>% 
  group_by(year.of.death,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_deaths = n_distinct(id),
            .groups = "drop") %>% 
  rename(year = year.of.death)
  
# group population by year, worker/manager, and age category
# filter to non-white people only 
# count the number of people in each strata
df_people_per_stratum = df_ipums_restricted_emp %>% 
  filter(white_non_white == "White") %>% 
  group_by(year,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_people = sum(perwt),
            .groups = "drop")
  
# get worker death rates by year and age group by joining to the number of people
# in worker strata and dividing deaths by people
df_worker_death_rates_crude = df_deaths_per_stratum %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  mutate(death_rate_crude_worker = sum_deaths / sum_people)

# calculate crude worker death rate overall by year for workers
df_worker_rates_crude_year = df_deaths_per_stratum %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  group_by(year) %>% 
  summarize(total_worker_deaths = sum(sum_deaths),
            total_workers = sum(sum_people),
            .groups = "drop") %>% 
  mutate(worker_crude_death_rate = total_worker_deaths / total_workers)
  
# start with number of people per stratum for managers only
# join to get the worker death rates per stratum
# calculate expected deaths as number of managers times worker death rates per stratum
# join to get actual number of manager deaths per stratum
df_manager_rates = df_people_per_stratum %>% 
  filter(worker_manager == "Manager") %>% 
  left_join(df_worker_death_rates_crude %>% 
              select(year, 
                     age_cat_every_5, 
                     death_rate_crude_worker),
            by = c("year", "age_cat_every_5")) %>% 
  mutate(expected_deaths = round(sum_people * death_rate_crude_worker, 2)) %>% 
  left_join(df_deaths_per_stratum %>% 
              filter(worker_manager == "Manager"),
            by = c("year", "worker_manager", "age_cat_every_5"))
  
# start with the manager expected and actual death counts by stratum 
# group into years and sum the expected and actual death counts by year
# divide actual by expected to get the adjusted mortality rate by year
# join back to worker crude rates by year 
df_manager_rates_adj_year = df_manager_rates %>% 
  group_by(year) %>% 
  summarize(total_expected = sum(expected_deaths),
            total_observed = sum(sum_deaths),
            .groups = "drop") %>% 
  mutate(manager_std_mortality_ratio = round(total_observed / total_expected, 2)) %>% 
  left_join(df_worker_rates_crude_year,
           by = "year") %>% 
  mutate(manager_adj_mortality_rate = manager_std_mortality_ratio * worker_crude_death_rate)
```


```{r}
#| output: true
df_manager_rates_adj_year %>% 
  mutate(worker_crude_death_rate = round(worker_crude_death_rate, 5),
         total_expected = round(total_expected,2)) %>% 
  relocate(year,
           total_worker_deaths,
           total_workers, 
           worker_crude_death_rate,
           total_expected,
           total_observed,
           manager_std_mortality_ratio,
           manager_adj_mortality_rate) %>% 
  dt_table() %>% 
  add_commas("total_workers")
```



### Age -- Non-White Workers (Reference) vs. White Managers
```{r}
# group deaths by year, worker/manager, and age category
# filter to non-white people only 
# count the number of deaths in each strata
df_deaths_per_stratum_NW = df_cohort1_restricted %>%
  filter(white_non_white == "Non-White") %>% 
  group_by(year.of.death,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_deaths = n_distinct(id),
            .groups = "drop") %>% 
  rename(year = year.of.death)
  
# group population by year, worker/manager, and age category
# filter to non-white people only 
# count the number of people in each strata
df_people_per_stratum_NW = df_ipums_restricted_emp %>% 
  filter(white_non_white == "Non-White") %>% 
  group_by(year,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_people = sum(perwt),
            .groups = "drop")

# group deaths by year, worker/manager, and age category
# filter to white people only 
# count the number of deaths in each strata
df_deaths_per_stratum_W = df_cohort1_restricted %>%
  filter(white_non_white == "White") %>% 
  group_by(year.of.death,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_deaths = n_distinct(id),
            .groups = "drop") %>% 
  rename(year = year.of.death)
  
# group population by year, worker/manager, and age category
# filter to white people only 
# count the number of people in each strata
df_people_per_stratum_W = df_ipums_restricted_emp %>% 
  filter(white_non_white == "White") %>% 
  group_by(year,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_people = sum(perwt),
            .groups = "drop")
  
# get Non-White worker death rates by year and age group by joining to the number of people
# in worker strata and dividing deaths by people
df_worker_death_rates_crude = df_deaths_per_stratum_NW %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum_NW,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  mutate(death_rate_crude_worker = sum_deaths / sum_people)

# calculate crude Non-White worker death rate overall by year for workers
df_worker_rates_crude_year = df_deaths_per_stratum_NW %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum_NW,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  group_by(year) %>% 
  summarize(total_worker_deaths = sum(sum_deaths),
            total_workers = sum(sum_people),
            .groups = "drop") %>% 
  mutate(worker_crude_death_rate = total_worker_deaths / total_workers)
  
# start with number of White people per stratum for managers only
# join to get the Non-White worker death rates per stratum
# calculate expected deaths as number of managers times worker death rates per stratum
# join to get actual number of White manager deaths per stratum
df_manager_rates = df_people_per_stratum_W %>% 
  filter(worker_manager == "Manager") %>% 
  left_join(df_worker_death_rates_crude %>% 
              select(year, 
                     age_cat_every_5, 
                     death_rate_crude_worker),
            by = c("year", "age_cat_every_5")) %>% 
  mutate(expected_deaths = round(sum_people * death_rate_crude_worker, 2)) %>% 
  left_join(df_deaths_per_stratum_W %>% 
              filter(worker_manager == "Manager"),
            by = c("year", "worker_manager", "age_cat_every_5"))
  
# start with the manager expected and actual death counts by stratum 
# group into years and sum the expected and actual death counts by year
# divide actual by expected to get the adjusted mortality rate by year
# join back to worker crude rates by year 
df_manager_rates_adj_year = df_manager_rates %>% 
  group_by(year) %>% 
  summarize(total_expected = sum(expected_deaths),
            total_observed = sum(sum_deaths),
            .groups = "drop") %>% 
  mutate(manager_std_mortality_ratio = round(total_observed / total_expected, 2)) %>% 
  left_join(df_worker_rates_crude_year,
           by = "year") %>% 
  mutate(manager_adj_mortality_rate = manager_std_mortality_ratio * worker_crude_death_rate)
```

```{r}
#| output: true
df_manager_rates_adj_year %>% 
  mutate(worker_crude_death_rate = round(worker_crude_death_rate, 5),
         total_expected = round(total_expected,2)) %>% 
  relocate(year,
           total_worker_deaths,
           total_workers, 
           worker_crude_death_rate,
           total_expected,
           total_observed,
           manager_std_mortality_ratio,
           manager_adj_mortality_rate) %>% 
  dt_table() %>% 
  add_commas("total_workers")
```


### Age -- White Workers (Reference) vs. Non-White Managers
```{r}
# group deaths by year, worker/manager, and age category
# filter to non-white people only 
# count the number of deaths in each strata
df_deaths_per_stratum_NW = df_cohort1_restricted %>%
  filter(white_non_white == "Non-White") %>% 
  group_by(year.of.death,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_deaths = n_distinct(id),
            .groups = "drop") %>% 
  rename(year = year.of.death)
  
# group population by year, worker/manager, and age category
# filter to non-white people only 
# count the number of people in each strata
df_people_per_stratum_NW = df_ipums_restricted_emp %>% 
  filter(white_non_white == "Non-White") %>% 
  group_by(year,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_people = sum(perwt),
            .groups = "drop")

# group deaths by year, worker/manager, and age category
# filter to white people only 
# count the number of deaths in each strata
df_deaths_per_stratum_W = df_cohort1_restricted %>%
  filter(white_non_white == "White") %>% 
  group_by(year.of.death,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_deaths = n_distinct(id),
            .groups = "drop") %>% 
  rename(year = year.of.death)
  
# group population by year, worker/manager, and age category
# filter to white people only 
# count the number of people in each strata
df_people_per_stratum_W = df_ipums_restricted_emp %>% 
  filter(white_non_white == "White") %>% 
  group_by(year,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_people = sum(perwt),
            .groups = "drop")
  
# get White worker death rates by year and age group by joining to the number of people
# in worker strata and dividing deaths by people
df_worker_death_rates_crude = df_deaths_per_stratum_W %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum_W,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  mutate(death_rate_crude_worker = sum_deaths / sum_people)

# calculate crude White worker death rate overall by year for workers
df_worker_rates_crude_year = df_deaths_per_stratum_W %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum_W,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  group_by(year) %>% 
  summarize(total_worker_deaths = sum(sum_deaths),
            total_workers = sum(sum_people),
            .groups = "drop") %>% 
  mutate(worker_crude_death_rate = total_worker_deaths / total_workers)
  
# start with number of White people per stratum for managers only
# join to get the Non-White worker death rates per stratum
# calculate expected deaths as number of managers times worker death rates per stratum
# join to get actual number of White manager deaths per stratum
df_manager_rates = df_people_per_stratum_NW %>% 
  filter(worker_manager == "Manager") %>% 
  left_join(df_worker_death_rates_crude %>% 
              select(year, 
                     age_cat_every_5, 
                     death_rate_crude_worker),
            by = c("year", "age_cat_every_5")) %>% 
  mutate(expected_deaths = round(sum_people * death_rate_crude_worker, 2)) %>% 
  left_join(df_deaths_per_stratum_NW %>% 
              filter(worker_manager == "Manager"),
            by = c("year", "worker_manager", "age_cat_every_5"))
  
# start with the manager expected and actual death counts by stratum 
# group into years and sum the expected and actual death counts by year
# divide actual by expected to get the adjusted mortality rate by year
# join back to worker crude rates by year 
df_manager_rates_adj_year = df_manager_rates %>% 
  group_by(year) %>% 
  summarize(total_expected = sum(expected_deaths),
            total_observed = sum(sum_deaths),
            .groups = "drop") %>% 
  mutate(manager_std_mortality_ratio = round(total_observed / total_expected, 2)) %>% 
  left_join(df_worker_rates_crude_year,
           by = "year") %>% 
  mutate(manager_adj_mortality_rate = manager_std_mortality_ratio * worker_crude_death_rate)
```

```{r}
#| output: true
df_manager_rates_adj_year %>% 
  mutate(worker_crude_death_rate = round(worker_crude_death_rate, 5),
         total_expected = round(total_expected,2)) %>% 
  relocate(year,
           total_worker_deaths,
           total_workers, 
           worker_crude_death_rate,
           total_expected,
           total_observed,
           manager_std_mortality_ratio,
           manager_adj_mortality_rate) %>% 
  dt_table() %>% 
  add_commas("total_workers")
```

### Age, Male Only-- Workers (Reference) Vs. Managers 
```{r}
# group deaths by year, worker/manager, and age category and count the number of 
# deaths in each strata
df_deaths_per_stratum = df_cohort1_restricted %>%
  filter(recode_sex == 'male') %>% 
  group_by(year.of.death,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_deaths = n_distinct(id),
            .groups = "drop") %>% 
  rename(year = year.of.death)
  
# group population by year, worker/manager, and age category and count the number
# of people in each strata
df_people_per_stratum = df_ipums_restricted_emp %>% 
  filter(sex == "Male") %>% 
  group_by(year,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_people = sum(perwt),
            .groups = "drop")
  
# get worker death rates by year and age group by joining to the number of people
# in worker strata and dividing deaths by people
df_worker_death_rates_crude = df_deaths_per_stratum %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  mutate(death_rate_crude_worker = sum_deaths / sum_people)

# calculate crude worker death rate overall by year for workers
df_worker_rates_crude_year = df_deaths_per_stratum %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  group_by(year) %>% 
  summarize(total_worker_deaths = sum(sum_deaths),
            total_workers = sum(sum_people),
            .groups = "drop") %>% 
  mutate(worker_crude_death_rate = total_worker_deaths / total_workers)
  
# start with number of people per stratum for managers only
# join to get the worker death rates per stratum
# calculate expected deaths as number of managers times worker death rates per stratum
# join to get actual number of manager deaths per stratum
df_manager_rates = df_people_per_stratum %>% 
  filter(worker_manager == "Manager") %>% 
  left_join(df_worker_death_rates_crude %>% 
              select(year, 
                     age_cat_every_5, 
                     death_rate_crude_worker),
            by = c("year", "age_cat_every_5")) %>% 
  mutate(expected_deaths = round(sum_people * death_rate_crude_worker, 2)) %>% 
  left_join(df_deaths_per_stratum %>% 
              filter(worker_manager == "Manager"),
            by = c("year", "worker_manager", "age_cat_every_5"))
  
# start with the manager expected and actual death counts by stratum 
# group into years and sum the expected and actual death counts by year
# divide actual by expected to get the adjusted mortality rate by year
# join back to worker crude rates by year 
df_manager_rates_adj_year = df_manager_rates %>% 
  group_by(year) %>% 
  summarize(total_expected = sum(expected_deaths),
            total_observed = sum(sum_deaths),
            .groups = "drop") %>% 
  mutate(manager_std_mortality_ratio = round(total_observed / total_expected, 2)) %>% 
  left_join(df_worker_rates_crude_year,
           by = "year") %>% 
  mutate(manager_adj_mortality_rate = manager_std_mortality_ratio * worker_crude_death_rate)
```


```{r}
#| output: true
df_manager_rates_adj_year %>% 
  mutate(worker_crude_death_rate = round(worker_crude_death_rate, 5)) %>% 
  relocate(year,
           total_worker_deaths,
           total_workers, 
           worker_crude_death_rate,
           total_expected,
           total_observed,
           manager_std_mortality_ratio,
           manager_adj_mortality_rate) %>% 
  dt_table() %>% 
  add_commas("total_workers")
```


### Age -- Non-White Workers (Reference) vs. White Workers
```{r}
# group deaths by year, worker/manager, and age category
# filter to non-white people only 
# count the number of deaths in each strata
df_deaths_per_stratum_NW = df_cohort1_restricted %>%
  filter(white_non_white == "Non-White") %>% 
  group_by(year.of.death,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_deaths = n_distinct(id),
            .groups = "drop") %>% 
  rename(year = year.of.death)
  
# group population by year, worker/manager, and age category
# filter to non-white people only 
# count the number of people in each strata
df_people_per_stratum_NW = df_ipums_restricted_emp %>% 
  filter(white_non_white == "Non-White") %>% 
  group_by(year,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_people = sum(perwt),
            .groups = "drop")

# group deaths by year, worker/manager, and age category
# filter to white people only 
# count the number of deaths in each strata
df_deaths_per_stratum_W = df_cohort1_restricted %>%
  filter(white_non_white == "White") %>% 
  group_by(year.of.death,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_deaths = n_distinct(id),
            .groups = "drop") %>% 
  rename(year = year.of.death)
  
# group population by year, worker/manager, and age category
# filter to white people only 
# count the number of people in each strata
df_people_per_stratum_W = df_ipums_restricted_emp %>% 
  filter(white_non_white == "White") %>% 
  group_by(year,
           worker_manager,
           age_cat_every_5) %>% 
  summarize(sum_people = sum(perwt),
            .groups = "drop")
  
# get Non-White worker death rates by year and age group by joining to the number of people
# in worker strata and dividing deaths by people
df_worker_death_rates_crude = df_deaths_per_stratum_NW %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum_NW,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  mutate(death_rate_crude_worker = sum_deaths / sum_people)

# calculate crude Non-White worker death rate overall by year for workers
df_worker_rates_crude_year = df_deaths_per_stratum_NW %>%  
  filter(worker_manager == "Worker") %>% 
  left_join(df_people_per_stratum_NW,
            by = c("year","worker_manager", "age_cat_every_5")) %>% 
  group_by(year) %>% 
  summarize(total_worker_deaths = sum(sum_deaths),
            total_workers = sum(sum_people),
            .groups = "drop") %>% 
  mutate(worker_crude_death_rate = total_worker_deaths / total_workers)
  
# start with number of White people per stratum for workers only
# join to get the Non-White worker death rates per stratum
# calculate expected deaths as number of managers times worker death rates per stratum
# join to get actual number of White worker deaths per stratum
df_manager_rates = df_people_per_stratum_W %>% 
  filter(worker_manager == "Worker") %>% 
  left_join(df_worker_death_rates_crude %>% 
              select(year, 
                     age_cat_every_5, 
                     death_rate_crude_worker),
            by = c("year", "age_cat_every_5")) %>% 
  mutate(expected_deaths = round(sum_people * death_rate_crude_worker, 2)) %>% 
  left_join(df_deaths_per_stratum_W %>% 
              filter(worker_manager == "Worker"),
            by = c("year", "worker_manager", "age_cat_every_5"))
  
# start with the manager expected and actual death counts by stratum 
# group into years and sum the expected and actual death counts by year
# divide actual by expected to get the adjusted mortality rate by year
# join back to worker crude rates by year 
df_manager_rates_adj_year = df_manager_rates %>% 
  group_by(year) %>% 
  summarize(total_expected = sum(expected_deaths),
            total_observed = sum(sum_deaths),
            .groups = "drop") %>% 
  mutate(std_mortality_ratio = round(total_observed / total_expected, 2)) %>% 
  left_join(df_worker_rates_crude_year,
           by = "year") %>% 
  mutate(adj_mortality_rate = std_mortality_ratio * worker_crude_death_rate)
```

```{r}
#| output: true
df_manager_rates_adj_year %>% 
  mutate(worker_crude_death_rate = round(worker_crude_death_rate, 5),
         total_expected = round(total_expected,2)) %>% 
  relocate(year,
           total_worker_deaths,
           total_workers, 
           worker_crude_death_rate,
           total_expected,
           total_observed,
           std_mortality_ratio,
           adj_mortality_rate) %>% 
  dt_table() %>% 
  add_commas("total_workers")
```



